---
title: "manuscript_public_code"
output: github_document
date: "2025-04-14"
---

This is an R-markdown to completely re-generate the figures from the ReVSeq Manuscript.

```{r}
#sourcing all packages
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(cowplot)
library(tidyr)
library(stringdist)
library(ggalluvial)
library(ggpubr)

library(ggtree)
library(treeio)
library(lubridate)
library(aplot)
library(scales)

library(ggrastr)
library(ape)
library(ggplotify)
```


```{r}
#sourcing all files 

#must upload virus abbreviation CSV

#overview 
source("Overview_circulating_viruses/fig_1_data_format.R")
source("Overview_circulating_viruses/fig_1_plot.R")

#PCR to seq comparison
source("Sequencing_results/fig_2_data_format.R")
source("Sequencing_results/fig_2_plot.R")

#co-infections
source("Sequencing_results/fig_3_data_format.R")
source("Sequencing_results/fig_3_plot.R")

#coverage plots
source("Sequencing_results/coverage_genome_data_format.R") 
source("Sequencing_results/coverage_genome_plot.R")

#tree plots
source("Phylogenetic_analysis/tree_plot.R")
source("Phylogenetic_analysis/make_map.R")

#supplementary wastewater
source("Supplements/fig_s2_data_format.R")
source("Supplements/fig_s2_plot.R")

#supplementary correlation CT-sequencing
source("Supplements/fig_s3_CT_data_format.R")
source("Supplements/fig_s3_CT_plot.R")

#supplementary quantifications mapped reads and depth of sequencing
source("Supplements/table_s3_quantification.R")


```

```{r}
hq_data<-read.csv("Data/data/hq_data.csv")
pcr_data <- read.csv("Data/data/pcr_data.csv")
detected_data <- read.csv("Data/data/detected_data.csv")
sentinella_data <- get_sentinella_data()
```


# Figure 1: seasonal overview

```{r, warning=FALSE}
substrains_to_highlight = ""
df_plt <- format_grid_pathogen_plot(hq_data, pcr=FALSE)
freq_data<-format_frequencies_plot(pcr_data, hq_data, detected_data ,substrain_to_highlight = substrains_to_highlight)

#substrain_to highlight must be a vector
make_figure_one(df_plt, freq_data, sentinella_data ,highlight = FALSE, save=TRUE, pcr=FALSE )


```
# Figure 2: PCR / sequencing comparison

```{r}
#all strains included in the PCR panel
pcr_panel_strains <- (pcr_data %>%  separate_rows(strains_PCR, panels_PCR, values_PCR, sep = ",\\s*") )$strains_PCR %>% unique()

formatted_pcr_to_detected<-format_pcr_to_detect_data(pcr_data, detected_data, pcr_panel_strains)
formatted_detected_to_hq<-format_detect_to_hq_data(detected_data, hq_data, pcr_panel_strains)

make_figure_comparison(formatted_pcr_to_detected, formatted_detected_to_hq, save=TRUE)

```



# Figure 3: Co-infections in high-quality sequences

```{r}
matrix_co <- format_matrix_co_infection_plot(hq_data)
percent_co <- format_percent_count_co_infection_plot(hq_data)

make_figure_3(matrix_co, percent_co, save=TRUE)
```
# Figure 4: Whole-genome coverage

```{r, warning=FALSE}
# RSV-A
#get data
virus_strain <- "Respiratory syncytial virus (type A)"
rsv_a_data <- hq_data %>% filter(grepl(virus_strain, substrain_name, fixed=TRUE))

df <- format_coverage_plot_data(rsv_a_data, virus_strain)
fig4a<-make_figure_coverage_genome(df, virus_strain, save=TRUE)
fig4a

# Parainfluenza-3
virus_strain <- "Human parainfluenza virus 3"
hpiv_3_data <- hq_data %>% filter(grepl(virus_strain, substrain_name, fixed=TRUE))

df <- format_coverage_plot_data(hpiv_3_data, virus_strain)
fig4b<-make_figure_coverage_genome(df, virus_strain, save=TRUE)
fig4b
```
```{r}
#plotting figures together

plots <- list( fig4a, fig4b)

plot_all<-plot_grid(plotlist = plots,
          #nrow = length(virus_strains),
          nrow = 2,
          rel_heights = c(1,1),
          labels= "AUTO",
          label_size= 12,
          label_y=1.02,
          label_x=0.01,
          align="v", 
          axis="lr"
          )

plot_all
ggsave("Figures/coverage_plots/F4.pdf",
       plot_all, 
       device = "pdf",
       dpi = 300, 
       units = "mm",
       width = 180,   # Max standard journal width
       height = 230   # Max standard height (leaving ~6cm for caption)
       )
```


# Figure 5: Phylogenetic trees

## RSV-A

## RSV-A full tree (Supplements)

```{r, warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_rsv_a_genome_2y_timetree.nexus"
metadata_file <- "Data/data/trees/nextstrain_rsv_a_genome_2y_metadata.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r}
#normal view (supplementary, full tree)
title ="RSV-A, 2-year"

# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("A.D.1", "A.D.1.5", "A.D.1.6", "A.D.2.1", "A.D.3", "A.D.3.1", "A.D.5.1", "A.D.5.2")

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         location_to_plot = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ location
        )
         
         )

results<-plot_tree_with_data(tr, meta, title  ,color_by="Clade", color_by_vertical = c( "Clade"), 
                    vertical_legend_on=c("Clade"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= FALSE, save=TRUE, tree_legend_on=F, date_format="%Y", tip_color = "database", full_page = TRUE )
results$plot
```
 
## RSV-A A.D clade, collapsed
 
```{r, warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_rsv_a_AD_clade.nexus"
metadata_file <- "Data/data/trees/nextstrain_rsv_a_AD_clade.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r}
# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("A.D.1", "A.D.1.5", "A.D.1.6", "A.D.2.1", "A.D.3", "A.D.3.1", "A.D.5.1", "A.D.5.2", "A.D.5")


tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         location_to_plot = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ location
        )
         
         )
```

```{r}

clades_to_collapse <- c("A.D.4", "A.D.5", "A.D.3", "A.D.1", "A.D.2.1", "A.D.2.2")
noisy_nodes <- c(2981,3995, 4076, 3991, 4002)
collapsed_tree_results <-tree_plot_collapsed(tr, meta, "Clade", "A.D", list_of_clades= clades_to_collapse, additional_node=noisy_nodes,
                    legend_position = c(0.05,0.8),
                    save=TRUE)
collapsed_tree_results
```
## RSV-A subclades:

### A.D.5
```{r, warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_rsv_a_AD5_clade.nexus"
metadata_file <- "Data/data/trees/nextstrain_rsv_a_AD5_clade.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r, warning=FALSE}
#normal view
title ="A.D.5"

# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("A.D.1", "A.D.1.5", "A.D.1.6", "A.D.2.1", "A.D.3", "A.D.3.1", "A.D.5.1", "A.D.5.2")

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         region = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern CH.",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ "Non-CH."
        )
         
         )

plot.a.d.5<-plot_tree_with_data(tr, meta, title  ,color_by="Clade", color_by_vertical = c("region"), 
                    vertical_legend_on=c("region"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= FALSE, save=FALSE, tree_legend_on=T, 
                    tree_legend_position = c(0.15,0.8),
                    date_format="%Y", tip_color = "database", full_page = FALSE,
                    small_mode = T)
plot.a.d.5$plot

```

### A.D.3

```{r, warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_rsv_a_AD3.nexus"
metadata_file <- "Data/data/trees/nextstrain_rsv_a_AD3.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r, warning=FALSE}
#normal view
title ="A.D.3"


# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("A.D.1", "A.D.1.5", "A.D.1.6", "A.D.2.1", "A.D.3", "A.D.3.1", "A.D.5.1", "A.D.5.2")

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         region = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern CH.",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ "Non-CH."
        )
         
         )

plot.a.d.3<-plot_tree_with_data(tr, meta, title  ,color_by="Clade", color_by_vertical = c("region"), 
                    vertical_legend_on=c("region"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= FALSE, save=FALSE, tree_legend_on=T, 
                    tree_legend_position = c(0.15,0.8),
                    date_format="%Y", tip_color = "database", full_page = FALSE,
                    small_mode = T)
plot.a.d.3$plot

```
### A.D.1

```{r, warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_rsv_a_AD1.nexus"
metadata_file <- "Data/data/trees/nextstrain_rsv_a_AD1.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r, warning=FALSE}
#normal view
title ="A.D.1"

# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("A.D.1", "A.D.1.5", "A.D.1.6", "A.D.2.1", "A.D.3", "A.D.3.1", "A.D.5.1", "A.D.5.2")

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         region = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern CH.",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ "Non-CH."
        )
         
         )

plot.a.d.1<-plot_tree_with_data(tr, meta, title  ,color_by="Clade", color_by_vertical = c("region"), 
                    vertical_legend_on=c("region"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= FALSE, save=FALSE, tree_legend_on=T, 
                    tree_legend_position = c(0.15,0.8),
                    date_format="%Y", tip_color = "database", full_page = FALSE,
                                                          small_mode=T )
plot.a.d.1$plot

```
### RSV-A Fig 5A

```{r}

map <- make_map(small_mode=TRUE)

fig5a1 <- plot_grid(
  map,
  collapsed_tree_results,
  ncol=1,
  rel_heights = c(0.2,1)
)


top_row <- plot_grid(
  fig5a1, plot.a.d.1$plot,
  nrow = 1,
  label_size = 12,
  rel_widths = c(1, 1.2)

)

# 2. Create Bottom Row (Equal Widths)
bottom_row <- plot_grid(
  plot.a.d.3$plot, plot.a.d.5$plot,
  nrow = 1,
  label_size = 12,
  rel_widths = c(1, 1) # Equal size
)

# 3. Stack
fig5a <- plot_grid(
  top_row,
  bottom_row,
  nrow = 2,
  align = "none",
  rel_heights = c(1, 1)
)

fig5a

ggsave("Figures/tree_plots/F5_a.pdf",
       fig5a,
       device = "pdf",
       dpi = 300,
       units = "mm",
       width = 180,   # Max standard journal width
       height = 130   # Max standard height (leaving ~6cm for caption)
       )
```

## RSV phenetic distances

```{r,warning=FALSE}
#phenetic distances between RSV-A and RSV-B

#load genetic distance trees
tree_file_a <- "Data/data/trees/div_nextstrain_rsv_a_genome_2y_tree.nexus"
tree_file_b <- "Data/data/trees/div_nextstrain_rsv_b_genome_2y_tree.nexus"
tr_a <- read.mega(tree_file_a)
tr_b <- read.mega(tree_file_b)
pat_mat_a<-cophenetic(tr_a@phylo)
pat_mat_b<-cophenetic(tr_b@phylo)


unique_distances_a <- pat_mat_a[lower.tri(pat_mat_a)]
unique_distances_b <- pat_mat_b[lower.tri(pat_mat_b)]

distance_df <- tibble(
  distance = c(unique_distances_a, unique_distances_b),
  source = c(rep("RSV-A", length(unique_distances_a)), rep("RSV-B", length(unique_distances_b)))
)

ks_result <- ks.test(unique_distances_a, unique_distances_b)

# --- 4. Print the results ---
print("Kolmogorov-Smirnov Test Results:")
print(ks_result)

distance_df %>% group_by(source) %>% summarise(median= median(distance), IQR=IQR(distance), q1= quantile(distance, 0.25), q3=quantile(distance, 0.75))
```

## HPIV-3

```{r,warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_HPIV_3_whole_timetree.nexus"
metadata_file <- "Data/data/trees/nextstrain_HPIV_3_whole_metadata.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

### HPIV-3 full tree (supplements)
```{r}
#normal view (supplementary, full tree)
title ="HPIV-3"

# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("C", "C3b")

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         location_to_plot = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ location
        )
         
         )

results<-plot_tree_with_data(tr, meta, title  ,color_by="Clade", color_by_vertical = c( "Clade", "region"), 
                    vertical_legend_on=c("Clade"=TRUE, "region"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= FALSE, save=T, tree_legend_on=T, date_format="%Y", tip_color = "database", full_page = TRUE )
results$plot
```


### HPIV-3 collapsed tree

```{r}
# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("C", "C3b")


tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         location_to_plot = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ location
        )
         
         )
```

```{r}

#clades_to_collapse <- c("A.D.4", "A.D.5", "A.D.3", "A.D.1", "A.D.2.1", "A.D.2.2")
#noisy_nodes <- c(2981,3995, 4076, 3991, 4002)
#collapsed_tree <-tree_plot_collapsed(tr, meta, "Clade", "A.D", list_of_clades= clades_to_collapse, additional_node=noisy_nodes,
#                    legend_position = c(0.05,0.8),
#                    save=TRUE)

groups_to_collapse <- c("C3b")
swiss_nodes <- c(1522, 992, 827, 845, 1165, 1044, 1345, 
                 1247, 1278, 862, 1386, 1063, 1094, 1487)
collapsed_tree <-tree_plot_collapsed(tr, meta, "Clade", "HPIV-3", list_of_clades= groups_to_collapse, additional_node=swiss_nodes,
                    legend_position = c(0.1,0.8),
                    save=T,
                    collapse_by_clade = T
                    )
collapsed_tree

```
```{r}
#find noisy nodes
tree_phylo<-as.phylo(tr)
node <- c("m2-jCKGBv","m2-Y4rZ7q") # top clade 1522
node <- c("m2-9jS4Ro","KF687358" )  #bottom, ancestral swiss 2nd node, 992

#pure noise:
node <- c("HM583801","OQ848522") #827
node <- c("PQ541044","OY757743") #845
node <- c("OQ981806","OQ848518") #1165 #second asian clade
node <- c("FJ455842", "MK679540") #1044
node <- c("PP411993","OY757731") #1345

#small branches from stem
node <- c("OQ848510","KY973595") #1247
node <- c("OR416978", "KY973590") #1278
node <- c("MW052041","KJ672574") #862
node <- c("MF034889","OQ848562")  #1386
node <- c("KJ672603","KF687350") #1063
node <- c("KJ672586","KJ672563") #1094
node <- c("MN306039", "MF118145") #1487

getMRCA(tree_phylo, node)
```

### HPIV-3 clade 1

```{r,warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_hpiv3_clade_1.nexus"
metadata_file <- "Data/data/trees/nextstrain_hpiv3_clade_1.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r, warning=FALSE}
#normal view
title ="clade 1"

# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("C", "C3b")

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         region = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern CH.",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ "Non-CH."
        )
         
         )

plot.hpiv3.clade.1 <-plot_tree_with_data(tr, meta, title  ,color_by="Clade", color_by_vertical = c("region"), 
                                         vertical_legend_on=c("region"=TRUE),
                                          vertical_plot_legend_columns = 1, 
                                          zoomed= FALSE, save=FALSE, tree_legend_on=T, 
                                          tree_legend_position = c(0.15,0.8),
                                          date_format="%Y", tip_color = "database", full_page = FALSE,
                                                                                small_mode=T )
plot.hpiv3.clade.1$plot

```
### HPIV-3 clade 2


```{r,warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_hpiv3_clade_2.nexus"
metadata_file <- "Data/data/trees/nextstrain_hpiv3_clade_2.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r, warning=FALSE}
#normal view
title ="clade 2"


# identifiy only clades that we are mentioning in the paper:
clades_to_color <- c("C", "C3b")

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_),
                             Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
                             
                             )
meta <- meta %>% 
  mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
         strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
         accession =strain,
         Clade = ifelse(clade_membership %in% clades_to_color, clade_membership, "other"), #simplifying clade encoding
         
         region = case_when(
          location == "TI"                          ~ "Ticino",
          location %in% c("GE", "VD", "VS")         ~ "Lake Geneva",
          location %in% c("BE", "FR", "SO", "NE", "JU") ~ "Espace Mittelland",
          location %in% c("BS", "BL", "AG")         ~ "Northwestern CH.",
          location == "ZH"                          ~ "Zurich",
          location %in% c("GL", "SH", "AI", "AR", "SG", "GR", "TG") ~ "Eastern CH.",
          location %in% c("LU", "UR", "SZ", "OW", "NW", "ZG")       ~ "Central CH.",
          TRUE                                      ~ "Non-CH."
        )
         
         )

plot.hpiv3.clade.2 <-plot_tree_with_data(tr, meta, title  ,color_by="Clade", color_by_vertical = c("region"), 
                                         vertical_legend_on=c("region"=TRUE),
                                          vertical_plot_legend_columns = 1, 
                                          zoomed= FALSE, save=FALSE, tree_legend_on=T, 
                                          tree_legend_position = c(0.15,0.8),
                                          date_format="%Y", tip_color = "database", full_page = FALSE,
                                                                                small_mode=T )
plot.hpiv3.clade.2$plot

```
### HPIV-3 Fig 5B

```{r}

map <- make_map(small_mode=TRUE, legend_right = F)



top_row <- plot_grid(
  collapsed_tree, map,
  nrow = 1,
  label_size = 12,
  rel_widths = c(1, 0.3)

)

# 2. Create Bottom Row (Equal Widths)
bottom_row <- plot_grid(
  plot.hpiv3.clade.1$plot, plot.hpiv3.clade.2$plot,
  nrow = 1,
  label_size = 12,
  rel_widths = c(1, 1) # Equal size
)

# 3. Stack
fig5b <- plot_grid(
  top_row,
  bottom_row,
  nrow = 2,
  align = "none",
  rel_heights = c(0.7, 1)
)

fig5b

ggsave("Figures/tree_plots/F5_b.pdf",
       fig5b,
       device = "pdf",
       dpi = 300,
       units = "mm",
       width = 180,   # Max standard journal width
       height = 100   # Max standard height (leaving ~6cm for caption)
       )
```




# Table S2 virus counts by each testing type
 
```{r}
df_pcr <-pcr_data %>% separate_longer_delim(cols = c(strains_PCR, panels_PCR, values_PCR), delim = ",") %>%
    # Remove leading/trailing whitespace
    mutate(strains_PCR = str_trim(strains_PCR),
           panels_PCR = str_trim(panels_PCR),
           values_PCR= str_trim(values_PCR)) %>% 
  mutate(
    panel_type = ifelse(grepl("resp", panels_PCR), "resp", ifelse(panels_PCR == "flua" | panels_PCR == "flub" | panels_PCR == "rspc", "p2", "p1"))
      ) %>% 
  filter(!is.na(strains_PCR)) %>% 
  dplyr::count(panel_type, strains_PCR)

df_detect <- detected_data %>% dplyr::count(strain_name)
df_hq <-hq_data %>% dplyr::count(strain_name)


```

# Table S3 quantification of mapped reads and depth of sequencing per strain and per PCR panel

```{r}
table_sx <-make_supplementary_table_quantification_sequencing(hq_data)
table_sx

```



# Figure S1, Figure S2: wastewater and sentinella correlations


```{r,warning=FALSE}
sentinella_df <- get_sentinella_data() 
wastewater_df <- get_wastewater_data()
pathogens <- sentinella_df$pathogen_name %>% unique() %>% unlist()

pcr_dat_supp <- pcr_data %>% 
  separate_longer_delim(cols = c(strains_PCR, panels_PCR, values_PCR), delim = ",") %>%
    # Remove leading/trailing whitespace
  mutate(strains_PCR = str_trim(strains_PCR),
           panels_PCR = str_trim(panels_PCR),
           values_PCR= str_trim(values_PCR)) %>% 
  rename(substrain_name = strains_PCR)


pcr_dat<-format_grid_pathogen_plot(pcr_dat_supp, pcr=TRUE) %>% drop_na()
detected_dat <- format_grid_pathogen_plot(detected_data, pcr=FALSE) %>% drop_na()
sequencing_data <- format_grid_pathogen_plot(hq_data, pcr=FALSE) %>% drop_na()


make_supplementary_figure_2(sentinella_df, pcr_dat, sequencing_data, detected_dat, wastewater_df , pathogens=pathogens ,save=TRUE)
```

# Figure S3: CT value / comparison to mapped reads and genome covered

```{r}

df <- format_ct_value_data(hq_data)
plot_ct_value_supplements(df, save=TRUE)

```


# Figure S4 A,B,C

```{r, warning=FALSE}
# RSV-B
#get data
virus_strain <- "Human Respiratory syncytial virus 9320 (type B)"
rsv_b_data <- hq_data %>% filter(grepl(virus_strain, substrain_name, fixed=TRUE))

df <- format_coverage_plot_data(rsv_b_data, virus_strain)
s3a<-make_figure_coverage_genome(df, virus_strain, save=TRUE)
s3a

# HMPV
#get data
virus_strain <- "metapneumovirus"
hmpv_data <- hq_data %>% filter(grepl(virus_strain, substrain_name, fixed=TRUE))

df <- format_coverage_plot_data(hmpv_data, virus_strain)
s3b<-make_figure_coverage_genome(df, virus_strain, save=TRUE)
s3b

```


```{r, warning=FALSE}
#A/H1N1
#get data
virus_strain <- "2015(H1N1)"
h1n1_data <- hq_data %>% filter(grepl(virus_strain, substrain_name, fixed=TRUE))

df <- format_coverage_plot_data(h1n1_data, virus_strain)
figs<-make_figure_coverage_genome(df, virus_strain, save=FALSE)


plots <- list( figs$h1n1_HA, figs$h1n1_NA)

plot_all<-plot_grid(plotlist = plots,
          #nrow = length(virus_strains),
          ncol = 2,
          rel_heights = c(1,1),
          align = "v",
          axis="l")

plot_all
#ggsave("Figures/coverage_plots/h1n1_HA_NA.pdf",
#       plot_all , 
#       dpi = 300, #dpi="screen", 
#       units="mm", 
#       width=180, #174, 
#       height= 120,#123
       #scale= 1.2
#       )
```

```{r}
plots <- list(s3a, s3b, plot_all)

top_row <- plot_grid(
  plots[[1]], plots[[2]], # s3a and s3b
  labels = c("A", "B"),
  label_size = 12,
  nrow = 1,
  align = "h",  # Align horizontally
  axis = "tb"   # Align Top and Bottom axes
)

# 2. Combine Top Row with the Bottom Plot (Figure C)
# We use ncol = 1 to stack them vertically
final_figure <- plot_grid(
  top_row,        # The combined A+B
  plots[[3]],     # The full-width C (your previous 'plot_all')
  labels = c("", "C"), # Top row is already labeled; we add C to the bottom
  label_size = 12,
  ncol = 1,       # Stack vertically
  rel_heights = c(1, 1) # Optional: Give the bottom plot a bit more space if it's complex
)

final_figure
ggsave("Figures/coverage_plots/s3.pdf",
       final_figure, 
       device = "pdf",
       dpi = 300, 
       units = "mm",
       width = 180,   # Max standard journal width
       height = 230   # Max standard height (leaving ~6cm for caption)
       )
```



# Figure S5: RSV-B Phylogenetic trees

```{r,warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_rsv_b_genome_2y_timetree.nexus"
metadata_file <- "Data/data/trees/nextstrain_rsv_b_genome_2y_metadata.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r,warning=FALSE}
#normal view
title ="RSV-B, 2-year"

tr@data <-tr@data %>% mutate( #database= ifelse(database=="ReVSeq", database, NA_character_),
                             location = ifelse(str_length(location)==2 & country=="Switzerland", location, NA_character_))
meta <- meta %>% mutate(location = ifelse(str_length(location)==2& country=="Switzerland", location, NA_character_),
                        strain = ifelse(is.na(database),strain, str_replace_all(strain, " ", "")),
                        accession =strain)


plot_tree_with_data(tr, meta, title  ,color_by="clade_membership", color_by_vertical = c("clade_membership", "region"), 
                    vertical_legend_on=c("clade_membership"=TRUE,"region"=TRUE),
                    vertical_plot_legend_columns = 2, 
                    zoomed= FALSE, save=TRUE, tree_legend_on=FALSE, date_format="%Y", tip_color = "database" )
```


# Figure S5: A/H1N1/HA Phylogenetic trees

```{r,warning=FALSE}
tree_file <- "Data/data/trees/nextstrain_h1n1_HA_timetree.nexus"
metadata_file <- "Data/data/trees/nextstrain_h1n1_HA_metadata.tsv"
tr <- read.mega(tree_file)
meta <- readr::read_tsv(metadata_file, show_col_types=FALSE)
```

```{r,warning=FALSE}
#normal view
title ="H1N1 HA"
plot_tree_with_data(tr, meta, title  ,color_by="clade_membership", color_by_vertical = c("clade_membership", "region"), 
                    vertical_legend_on=c("clade_membership"=TRUE,"region"=FALSE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= FALSE, save=TRUE, tree_legend_on=FALSE, date_format="%Y", tip_color = "database" )




#zoomed in 
title ="H1N1 HA, clade 5a.2a"
plot_tree_with_data(tr, meta, title,  color_by="clade_membership", color_by_vertical = c("location", "country"), 
                    vertical_legend_on=c("country"=FALSE,"location"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= TRUE,tips_to_zoom = c("m2-4SnyCD", "m2-sFFUfJ"), axis_breaks = 30, 
                    save=TRUE, tree_legend_on=FALSE, date_format="%Y-%m", tip_color = "database" )


title ="H1N1 HA, clade 5a.2a.1"
plot_tree_with_data(tr, meta, title,  color_by="clade_membership", color_by_vertical = c("location", "country"), 
                    vertical_legend_on=c("country"=FALSE,"location"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= TRUE,tips_to_zoom = c("m2-3qDrEd", "m2-qeHUBn"), axis_breaks = 30, 
                    save=TRUE, tree_legend_on=FALSE, date_format="%Y-%m",tip_color = "database"  )

title ="variant expanding"
plot_tree_with_data(tr, meta, title,  color_by="clade_membership", color_by_vertical =c("location", "country"), 
                    vertical_legend_on=c("country"=FALSE,"location"=TRUE),
                    vertical_plot_legend_columns = 1, 
                    zoomed= TRUE,tips_to_zoom = c("m2-5cJWyf", "m2-kNU7js"), axis_breaks = 30, 
                    save=TRUE, tree_legend_on=FALSE, date_format="%Y-%m",tip_color = "database"  )
```




